---
title: "Analysis pass/fail"
output: html_document
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ballot Complexity and Ballot Success (Pass/Fail)

The CEDA codebook defines OUTCOME as 1 = Pass and 2 = Fail, while PASSFAIL gives the more detailed threshold-specific result.


## General Findings 

```{r}
library(tidyverse)
library(scales)
library(readxl)
```


### Load and Merge Datasets 

```{r}

setwd("/Users/mayajohnson/Desktop")

# Load CEDA measures dataset from 2024 
measures_2024 <- read_csv("Measures2024.csv", show_col_types = FALSE)

# Load voter registration info from CA in 2024 
registration <- read_excel("voter_registration_full.xlsx")

registration <- registration %>%
  mutate(county = str_to_upper(county))

# Add the total registered voter count per county to the measures dataset 
measures_2024 <- measures_2024 %>%
  mutate(CNTYNAME = str_to_upper(CNTYNAME)) %>%
  left_join(registration %>% select(county, total_registered),
            by = c("CNTYNAME" = "county"))

summary(measures_2024$total_registered)

# Double-check the merge worked 
measures_2024 %>%
  select(CNTYNAME, total_registered) %>%
  distinct() %>%
  head()

# Add a column in the measures dataset for the turnout percentage 
measures_2024 <- measures_2024 %>%
  mutate(
    turnout_pct = TOTAL / total_registered
  )
```

```{r}

measures_2024 <- measures_2024 %>%
  mutate(
    turnout = TOTAL,
    word_count = str_count(BALQUEST, "\\S+"),
    outcome_label = case_when(
      outcome == 1 ~ "Pass",
      outcome == 2 ~ "Fail",
      TRUE ~ NA_character_
    ),
    outcome = factor(outcome, levels = c("Pass", "Fail")),
    req_label = case_when(
      REQ == "M" ~ "Majority",
      REQ == "F" ~ "55%",
      REQ == "T" ~ "Two-thirds",
      TRUE ~ REQ
    ),
    jurisdiction = case_when(
      JUR == 1 ~ "County",
      JUR == 2 ~ "City",
      JUR == 3 ~ "School District",
      TRUE ~ "Other"
    )
  )
```

## Plot general pass vs. fail 

```{r}

measures_2024 %>%
  filter(!is.na(outcome_label)) %>%
  ggplot(aes(x = outcome_label, fill = outcome_label)) +
  geom_bar(show.legend = FALSE) +
  labs(
    title = "2024 California Local Ballot Measures: Pass vs. Fail",
    x = "Outcome",
    y = "Number of Measures"
  ) +
  theme_minimal()

```

## Plot Passing vs. Requirement Threshold to Pass  
```{r}
measures_2024 %>%
  filter(!is.na(outcome_label), !is.na(req_label)) %>%
  group_by(req_label) %>%
  summarise(
    total_measures = n(),
    passed = sum(outcome_label == "Pass"),
    passage_rate = passed / total_measures,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = req_label, fill = req_label, y = passage_rate)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Passage Rate by Required Approval Threshold",
    x = "Required Threshold",
    y = "Passage Rate"
  ) +
  theme_minimal()
```

## Plot Passage Rate by Jurisdiction Type 
```{r}

measures_2024 %>%
  filter(!is.na(outcome_label), !is.na(jurisdiction)) %>%
  group_by(jurisdiction) %>%
  summarise(
    n = n(),
    passage_rate = mean(outcome_label == "Pass"),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = jurisdiction, y = passage_rate)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Passage Rate by Jurisdiction Type",
    x = "Jurisdiction",
    y = "Passage Rate"
  ) +
  theme_minimal()

```


## Ballot question length by Pass/Fail outcome 
```{r}
measures_2024 %>%
  filter(!is.na(outcome_label), !is.na(word_count)) %>%
  ggplot(aes(x = outcome_label, y = word_count, fill = outcome_label)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Ballot Question Length by Outcome",
    x = "Outcome",
    y = "Word Count"
  ) +
  theme_minimal()

ggplot(measures_2024, aes(x = outcome_label, y = word_count, fill = outcome_label)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~ jurisdiction) +
  labs(
    title = "Ballot Question Length by Outcome, by Jurisdiction",
    x = "Outcome",
    y = "Word Count"
  ) +
  theme_minimal()
```



```{r}
measures_2024 %>%
  filter(!is.na(outcome_label), !is.na(turnout_pct)) %>%
  ggplot(aes(x = outcome_label, y = turnout_pct, fill = outcome_label)) +
  geom_boxplot(show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Measure Turnout by Outcome",
    x = "Outcome",
    y = "Total Votes"
  ) +
  theme_minimal()
```



## Average number of Words in Ballot Question vs. Turnout Rate 
```{r}
measures_2024 <- measures_2024 %>%
  mutate(
    turnout_pct = TOTAL / total_registered,
    word_count = str_count(BALQUEST, "\\S+")
  )

avg_words <- measures_2024 %>%
  filter(!is.na(CNTYNAME), !is.na(DATE), !is.na(turnout_pct), !is.na(word_count)) %>%
  group_by(CNTYNAME, DATE) %>%
  summarise(
    avg_word_count = mean(word_count, na.rm = TRUE),
    avg_turnout_pct = mean(turnout_pct, na.rm = TRUE),
    .groups = "drop"
  )

# Filter for one extreme outlier above 80
avg_words_filtered <- avg_words %>%
  filter(avg_word_count <= 80)

ggplot(avg_words_filtered, aes(x = avg_word_count, y = avg_turnout_pct)) +
  geom_point() +
  geom_smooth() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Average Ballot Question Length and Turnout Rate",
    x = "Average Number of Words in Ballot Questions",
    y = "Average Turnout (% of Registered Voters)"
  ) +
  theme_minimal()
```


```{r}
county_choice <- NULL
place_choice  <- NULL
date_choice   <- NULL

plot3_df <- measures_2024

if (!is.null(county_choice)) {
  plot3_df <- plot3_df %>% filter(CNTYNAME == county_choice)
}

if (!is.null(place_choice)) {
  plot3_df <- plot3_df %>% filter(PLACE == place_choice)
}

if (!is.null(date_choice)) {
  plot3_df <- plot3_df %>% filter(DATE == date_choice)
}

plot3_df <- plot3_df %>%
  filter(!is.na(turnout_pct), !is.na(LTR)) %>%
  filter(str_detect(LTR, "^[A-Z]$")) %>%
  mutate(ballot_position = factor(LTR, levels = LETTERS))

avg_votes_by_position <- plot3_df %>%
  group_by(ballot_position) %>%
  summarise(avg_votes = mean(turnout_pct, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(ballot_position))

ggplot(avg_votes_by_position, aes(x = ballot_position, y = avg_votes)) +
  geom_col() +
  labs(
    title = "Ballot Position and Average Turnout Rate",
    subtitle = "Average turnout rate by ballot position (A-Z) across all California Counties 2024",
    x = "Ballot Position",
    y = "Average Turnout Rate"
  ) +
  theme_minimal()

```

# County-Specific Case Studies 

## Los Angeles and Alameda County 
```{r}



```


## Marin County and Lassen County 
```{r}
marin_lassen <- measures_2024 %>%
  filter(CNTYNAME %in% c("MARIN", "LOS ANGELES"),
         JUR == 1) %>%   # ✅ keep only county-level measures
  mutate(
    turnout_pct = TOTAL / total_registered
  ) %>%
  filter(!is.na(turnout_pct), !is.na(word_count))

ggplot(marin_lassen, aes(x = word_count, y = turnout_pct, color = CNTYNAME)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Ballot Complexity and Turnout: Marin vs. Lassen Counties",
    x = "Ballot Question Length (Word Count)",
    y = "Turnout (% of Registered Voters)",
    color = "County"
  ) +
  theme_minimal()
```

```{r}
measures_2024 %>%
filter(CNTYNAME == "MARIN") %>%
select(total_registered) %>%
distinct()

measures_2024 %>%
  filter(CNTYNAME == "MARIN") %>%
  select(TOTAL) %>%
  summary()

measures_2024 %>%
  filter(CNTYNAME == "MARIN") %>%
  mutate(turnout_pct = TOTAL / total_registered) %>%
  select(TOTAL, total_registered, turnout_pct) %>%
  head(10)
```